RFC: Virtual Object Layer

### 2014-09-16 RFCs/HDF5/VOL/RFC/RFC_VOL.tex
more updates.
Former-commit-id: f993d3f92b5d8efb931abe1220cba82e90205fa4

% Title, author, etc
\title{RFC: Virtual Object Layer}
\author{Mohamad Chaarawi}
\author{Quincey Koziol}
\date{September 16, 2014}
\rfcversion{2014-09-16.v3}
\revision{January 9, 2012}{First Draft circulated internally.}
\revision{March 9, 2012}{Update and circulate internally and externally.}
\revision{August 28, 2014}{Update with new framework changes.}

@@

The VOL public structure separates Data Model operations from Service operations. Data Model operations are those that operate on files, groups, dataset, etc… Those operations are furthermore grouped in their respective ``class'' structure. For example the file class consists of function pointers for plugins to implement operations on the file itself (H5F API operations). Service operations are those that provide services for users that are not related to the data model specifically. Asynchronous operations, for example, are a service that most plugins can implement, so we add a class for it in the VOL structure. If a service becomes generic enough and common between many plugins, a class for it should be added in the VOL structure. However, many plugins can/will provide services that are not shared by other plugins. A good way to support these services is through an optional callback in the VOL structure which can be a hook from the API to the plugin that provides those services, passing any necessary arguments needed without the HDF5 library having to worry about supporting that service. A similar API operation to allow users to use that service will be added. This API call would be similar to an ``ioctl'' call where any kind of operation can be supported and passed down to the plugin that has enough knowledge from the user to interpret the type of the operation. 
The VOL User Guide will discuss the VOL structure in much more detail and provide examples on how to support/implement the classes/callbacks.

@@

	/* call the corresponding callback for H5Dcreate */
	result = H5VL_dataset_create (…); /* will route the operation to the plugin at the intermediate VOL layer */
	/* return result to user (yes the dataset is created, or no here is the error */
	return result;
}
\end{lstlisting}
}

@@

This would indicate that the MDS object layer plugin is to be used. All accesses to objects within the file opened using that FAPL would be routed through the MDS plugin.
In addition to plugins for the VOL class that will be added in the HDF5 library, any developer may write their own customizable plugin and tell the HDF5 library to use it at runtime. The plugin should be registered using:
{\lstsetc
\begin{lstlisting}
hid_t H5VLregister(H5VOL_class_t *cls)
\end{lstlisting}
}
The {\tt cls} parameter is a structure containing function pointers to the routines that implement all the callback routines the VOL class specifies.

@@

Another possible VOL plugin could be a statistics plugin that just gathers information on HDF5 API calls and records statistics associated with the number of calls to a specific API functions and corresponding parameters. This plugin would be very useful for profiling purposes. The statistics plugin would be stacked on top of another VOL plugin that actually performs the required access to the file.
\clearpage
\section{Implementation}
The VOL class structure consists of data structure(s) that contain configuration variables that are set by each plugin and a collection of function pointers that maps all HDF5 API operations that access data on disk to the plugin’s corresponding implementations. Internally, each plugin is free to implement the underlying storage of objects in any way desired. The main goal is that the user still gets the same HDF5 API and data access model no matter which plugin is used. This section highlights some implementation details and the preliminary API for the VOL.

@@

One could argue that the interface can be simplified by having, for example, a common open routine for all objects (like the {\tt H5Oopen}) routine. We found however that having  a common interface would be detrimental to performance in certain plugin implementations, since it would require querying the VOL plugin twice, once for looking up the object location and once for opening the object. Having an optional callback for every object and a generic one would solve this issue. The VOL layer would check if the specific object callback is implemented in the VOL plugin, and if it is not, it would fall back to the generic object callback routine. This way we don’t sacrifice performance for generality.

@@

@@

Comparing two objects that belong to different types of HDF5 containers require processing independent of the VOL plugins. One way to accomplish this is to add utility routines to the VOL class to retrieve certain characteristics about an object for comparison. For example, datasets would need utility routines to retrieve the dataset name, dimensions, type, space, and elements. Once those values are retrieved for each object from each VOL plugin, the comparison operation could return the required result. The copy operation would require the same level of processing on the source object. Then using the fetched attributes, the same object is created using a create callback of the destination object’s VOL plugin.

>>>>

% Title, author, etc
\title{RFC: Virtual Object Layer}
\author{Mohamad Chaarawi}
\author{Quincey Koziol}
\date{September 16, 2014}
\rfcversion{2014-09-16.v3}
\revision{January 9, 2012}{First Draft circulated internally.}
\revision{March 9, 2012}{Update and circulate internally and externally.}
\revision{August 28, 2014}{Update with new framework changes.}
\revision{September 16, 2014}{Update with new features.}

@@

The VOL public structure separates Data Model operations from Service operations. Data Model operations are those that operate on files, groups, dataset, etc... Those operations are furthermore grouped in their respective ``class'' structure. For example the file class consists of function pointers for plugins to implement operations on the file itself (H5F API operations). Service operations are those that provide services for users that are not related to the data model specifically. Asynchronous operations, for example, are a service that most plugins can implement, so we add a class for it in the VOL structure. If a service becomes generic enough and common between many plugins, a class for it should be added in the VOL structure. However, many plugins can/will provide services that are not shared by other plugins. A good way to support these services is through an optional callback in the VOL structure which can be a hook from the API to the plugin that provides those services, passing any necessary arguments needed without the HDF5 library having to worry about supporting that service. A similar API operation to allow users to use that service will be added. This API call would be similar to an ``ioctl'' call where any kind of operation can be supported and passed down to the plugin that has enough knowledge from the user to interpret the type of the operation. 
The VOL User Guide will discuss the VOL structure in much more detail and provide examples on how to support/implement the classes/callbacks.

@@

	/* call the corresponding callback for H5Dcreate */
	result = H5VL_dataset_create (...); /* will route the operation to the plugin at the intermediate VOL layer */
	/* return result to user (yes the dataset is created, or no here is the error */
	return result;
}
\end{lstlisting}
}

@@

This would indicate that the MDS object layer plugin is to be used. All accesses to objects within the file opened using that FAPL would be routed through the MDS plugin.
In addition to plugins for the VOL class that will be added in the HDF5 library, any developer may write their own customizable plugin and tell the HDF5 library to use it at runtime. The plugin should be registered using:
{\lstsetc
\begin{lstlisting}
hid_t H5VLregister(H5VOL_class_t *cls)
\end{lstlisting}
}
The {\tt cls} parameter is a structure containing function pointers to the routines that implement all the callback routines the VOL class specifies. For more information on creating and using VOL plugins, please refer to the VOL user guide.

@@

Another possible VOL plugin could be a statistics plugin that just gathers information on HDF5 API calls and records statistics associated with the number of calls to a specific API functions and corresponding parameters. This plugin would be very useful for profiling purposes. The statistics plugin would be stacked on top of another VOL plugin that actually performs the required access to the file.
\clearpage
\section{Implementation}
The VOL class structure consists of data structure(s) that contain configuration variables that are set by each plugin and a collection of function pointers that maps all HDF5 API operations that access data on disk to the plugin's corresponding implementations. Internally, each plugin is free to implement the underlying storage of objects in any way desired. The main goal is that the user still gets the same HDF5 API and data access model no matter which plugin is used. This section highlights some implementation and feature details.

@@

One could argue that the interface can be simplified by having, for example, a common open routine for all objects (like the {\tt H5Oopen}) routine. We found however that having  a common interface would be detrimental to performance in certain plugin implementations, since it would require querying the VOL plugin twice, once for looking up the object location and once for opening the object. Having an optional callback for every object and a generic one would solve this issue. The VOL layer would check if the specific object callback is implemented in the VOL plugin, and if it is not, it would fall back to the generic object callback routine. This way we do not sacrifice performance for generality.

@@

\subsection{Internal vs. External Plugins}
Developers can implement their VOL plugin as a plugin internal to the HDF5 library or an external plugin. Internal plugins are ones that rely heavily on internal and private HDF5 functionality like the native and metadata server plugins. They are required to ship with the HDF5 library to be used by applications. External plugins on the other hand are ones that are implemented outside of the HDF5 library and applications link with to use. Those plugins do not require using private HDF5 functionality. They can however use public HDF5 features.

External plugins can be provided by the applications themselves or as a third party library, i.e. not provided by the HDF5 library nor the application. Third party plugins should be installed on the system to be used as a shared library or a DLL. HDF5 will attempt to load the plugin if the application requests its use. This is similar to Filter plugins~\cite{filters}. Refer to the VOL user guide for more detailed information on how to develop and use VOL plugins.

@@

Comparing two objects that belong to different types of HDF5 containers require processing independent of the VOL plugins. One way to accomplish this is to add utility routines to the VOL class to retrieve certain characteristics about an object for comparison. For example, datasets would need utility routines to retrieve the dataset name, dimensions, type, space, and elements. Once those values are retrieved for each object from each VOL plugin, the comparison operation could return the required result. The copy operation would require the same level of processing on the source object. Then using the fetched attributes, the same object is created using a create callback of the destination object's VOL plugin.
